Lead distance under a pickoff limit in Major League Baseball:\\A sequential game model

### 2025-12-23 articles/arxiv/paper.tex
final(?) read throughs

\usepackage{amsmath, amssymb, amsthm, amsfonts, color, enumerate, framed, graphicx, mathrsfs, mathtools, mdframed, subcaption, tabularx, theoremref, titlesec, verbatim, xfrac, soul, mathtools, wrapfig, tikz, float}

@@

\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}
\newcommand{\defeq}{\stackrel{\text{def}}{=}}
\newcommand{\abs}[1]{\left\lvert#1\right\rvert}

\title{
  Lead distance under a pickoff limit in Major League Baseball:\\A sequential game model
}

@@

  \begin{abstract}
    Recently, Major League Baseball (MLB)  limited  pitcher to three pickoff attempts, creating a cat-and-mouse game between pitcher and runner. Each failed  attempt adds pressure on the pitcher to avoid using another, and the runner can intensify this pressure by extending their leadoff toward the next base. We model this dynamic as a  two-player zero-sum sequential game in which the runner first chooses a lead distance, and then the pitcher chooses whether to attempt a pickoff. We establish optimality characterizations for the game and present variants of value iteration and policy iteration to solve the game. Using  lead distance data, we estimate generalized linear mixed-effects models for pickoff and stolen base outcome probabilities given lead distance, context, and player skill. We compute the game-theoretic equilibria under the two-player model, as well as the optimal runner policy under a simplified one-player Markov decision process model. In the one-player setting, our results establish an actionable rule of thumb: the Two-Foot Rule, which recommends that a runner increase their lead by two feet after each pickoff attempt.
  \end{abstract}

@@

    On May 17, 2025, the Chicago Cubs hosted the Chicago White Sox in the second game of their Crosstown Series. In the bottom of the first inning, White Sox pitcher Sean Burke faced Cubs batter Michael Busch with two outs and Cubs runner Kyle Tucker on first base. Before the 1-1 pitch, Burke pivoted and threw to first base behind Tucker, who retreated safely before a tag. Tucker had 10 stolen bases in 10 attempts on the young season, leading Major League Baseball (MLB) among base stealers who had not been caught \parencite{fangraphs_major_2025}. On the White Sox broadcast, color commentator Steve Stone explained to the audience, ``The only problem is now there's two throws over there, and Tucker's gonna get a much bigger lead.'' On the next pitch, Tucker stole his 11$^{th}$ base of the season, the first Burke had allowed. ``I don't think anybody in the room could have thrown him out on that play,'' said Stone.

    For background, the stolen base is an exciting play in which a runner takes off on the pitch, attempting to reach the next base before the catcher can throw the ball to the base for a fielder to tag the runner out \parencite{major_league_baseball_baseball_2025}. Whether attempting a stolen base or not, runners typically take a {\it lead} from their base, commonly positioning themselves 8--12 feet away from their base, in the direction of the next base. The longer the lead, the shorter the distance needed to run to reach the next base. The pitcher's primary deterrent against long leads is a {\it pickoff attempt}, in which they throw the ball to a fielder at the runner's base, attempting a tag before the runner returns safely to the base.

    MLB stolen base totals surged in the 1980s and 1990s but then waned in the 2000s and 2010s as teams cut back on the play, which generally requires a high success rate in order to have a positive impact on run scoring \parencite{tango_book_2007}. Before 2023, pickoff attempts could cause indefinite delays in the progress of a game because unsuccessful pickoff attempts do not advance the state of the game. Acknowledging this and other trends, MLB introduced several rule changes starting in 2023, with the goal of making the game more exciting \parencite{castrovince_pitch_2023}. With these changes, a pitch clock limits the time between pitches, and there are restrictions on infielder positioning, curtailing defensive shifts. Two changes are particularly relevant to stolen base attempts: Bases are now wider (reducing the distance between first and second base by 4.5 inches), and there are now limitations on pickoff attempts, or more formally, {\it disengagements}.

@@

    {\color{blue}{Section \ref{sec:sg-model} uses a two-payer zero-sum sequential game to model the evolving interaction between the runner and the pitcher.  Our model captures all the vital aspects of the play while retaining analytical and computational tractability. Our model is closely related to the class of stochastic/dynamic Stackelberg games \parencite{goktas2022zero,li2017review,lopez2022stationary,vasal2020stochastic,vorobeychik2012computing}, which combines the elements of Stackelberg and stochastic games. The main difference between our infinite-horizon model and the literature on stochastic Stackelberg games is that we do not discount the rewards. The rationale for this modeling choice is that we choose our optimization objective to be expected runs in the inning, which by definition includes no discount. We establish a maximin Bellman equation for our sequential game and propose a variant of value iteration and policy iteration to determine the optimal strategies for runner and the pitcher.}}

    In addition to developing a model which captures the important features of the pitcher-runner game, we analyze the model using recently available real-world MLB data. The most innovative component of our implementation is the construction of transition probabilities. The outcome of each play is influenced by factors outside of our state variable, such as the sprint speed of the runner and the arm strength of the catcher. As a result, constructing transition probabilities using empirical conditional estimates (which use only the information from the current state and the actions of the player) is likely to be less accurate compared with a method that incorporates the player-specific factors.

    Section \ref{sec:transition-probability-model} uses a two-step procedure that includes the player-specific factors to estimate the transition probabilities. In the first step, we use a mixed-effects logistic regression model that calculates the probability of various runner outcomes taking into account the actions of the players and the player-specific information. Next, we construct empirical estimates of the next state conditioned upon the current state, runner outcomes, and the actions of the player. Finally, we determine the transition probability via these two probabilities.

    Section \ref{sec:results} presents a numerical study on recent real-world data which illustrates that, relative to the optimal policy, observed pitcher and runner decisions are conservative. Pitchers attempt fewer pickoffs than optimal, and runners take shorter leads than optimal. Relative to observed behavior, optimal play would favor the offense slightly. Moreover, when restricting out model to a single-player setup which allows agency only for the runner, the results reveal an actionable recommendation: The Two-Foot Rule, which recommends that a runner increase their lead by two feet after each pickoff attempt.

@@

    Several studies in the sports economics literature have examined the extent to which professional athletes play minimax strategy. \textcite{palacios-huerta_professionals_2003} found evidence consistent with minimax equilibrium behavior in the mixed strategy game of the penalty kick, in which the kicker chooses to shoot the ball left or right and the goalkeeper simultaneously chooses to dive left or right for the save. \textcite{palacios-huerta_experientia_2008} followed this with a laboratory experiment featuring professional soccer players and college students simulating the penalty kick decision using cards, showing that the professionals played minimax but the students did not. On the other hand, \textcite{kovash_professionals_2009} found evidence inconsistent with minimax equilibrium behavior in the National Football League (teams called more run plays and fewer pass plays than theoretically optimal) and in MLB (pitchers threw more fastballs than theoretically optimal).

    The history of quantitative analysis in baseball has deep roots in operations research (e.g., \cite{lindsey_investigation_1963}; \cite{freeze_analysis_1974}; \cite{pankin_evaluating_1978}; \cite{russell_devising_1994}; \cite{bukiet_markov_1997}). The Markov process is particularly well suited for baseball, where the game's highly structured progression lends it well to an intuitive state space model \parencite{sokol_intuitive_2004}. \textcite{hirotsu_markov_2003} used a Markov model to inform substitution (pinch hitting) decision-making in baseball. The Markov process is also popular across sport analytics. \textcite{hirotsu_using_2002} used a four-state Markov model for soccer to inform substitution decision-making based on how substitutions impact transition probabilities between states. Recently, \textcite{chan_points_2021} used a Markov model to evaluate team performance in football. More specifically, the MDP has become particularly popular model in recent years for analyzing decision-making in sports, with applications in tennis \parencite{nadimpalli_when_2013}, baseball \parencite{hirotsu_using_2019}, basketball \parencite{sandholtz_markov_2020}, darts \parencite{baird_optimising_2020} and football \parencite{biro_reinforcement_2022}.

    An intermediate result of the present work is a probability model for stolen base attempts and successes (as well as an analysis how those probabilities changed with the 2023 MLB rule changes), a topic which has not received much attention. \textcite{loughin_assessing_2008} estimated linear mixed-effects models for stolen base attempts and successes, using random effects for pitcher and catcher (not the runner, notably) and fixed effects for balls, strikes, outs, score, venue and pitcher hand. They showed that the spread of pitcher effects is slightly greater than the spread of catcher effects. That analysis did not use data on runner sprint speed and catcher arm strength. More recently, \textcite{stanley_modeling_2023} modeled stolen base success probability using both logistic regression and a random forest, including runner sprint speed and catcher arm strength as features but excluding player identities. That analysis was limited to data prior to the 2023 MLB rule changes that incentivized base stealing.

    \section{The two-player sequential game model}
    \label{sec:sg-model}
    In this section, we present our two-player zero-sum sequential game model under the pickoff limit rule. Though our sequential game model describes only a portion of an inning, it captures all the vital aspects that enable the generation of valuable and actionable insights for the runner and pitcher. Before describing the specifics of our model, we note that our model assumes a representative pitcher and a representative runner for the whole inning, rather than capturing full lineups and substitutions of players with varying skills. Without this assumption, we will end up with a nonstationary (infinite-horizon) two-player sequential game model, which appears to be significantly more challenging to analyze when compared with the stationary version. The four components of our sequential game are as follows.

    {\bf 1. State space.} We use $s=(b,c,d,o) \in {\cal X}$ to capture a play, where     
    % The mid-inning game state $s \in \mathcal{M}$ is given by $s = (b, c, d, o)$, where:
    \begin{align*}
      b &\in \{0, 1\}^3 \mbox{ represents which bases are occupied by runners};\\
      c &\in \{0, 1, 2, 3\} \times \{0, 1, 2\} \mbox{ represents the ball-strike count};\\
      d &\in \{0, 1, 2\} \mbox{ represents the number of disengagements already used}; and\\
      o &\in \{0, 1, 2\} \mbox{ represents the number of outs}.
    \end{align*}
    The number of elements in ${\cal X}$ equals $2^3 \times (4 \times 3) \times 3 \times 3 = 864$. We use $\mathcal P$ to denote the set of four penultimate states and $\Delta$ to denote the terminal end-of-inning state. The purpose of the penultimate states is to keep track of the number of runs that are scored on the transition to the terminal state. A penultimate state $s \in \mathcal{P} = \{0, 1, 2, 3\}$ represents the number of runs scored on the final play of the inning. We use $\mathcal{S} \equiv \mathcal{X} \cup \mathcal{P} \cup \{\Delta\}$ to denote the set of all states, and $S_t$ to denote the state at play $t$ in a (fixed) inning. Traditionally, the state is defined by the bases occupied and the number of outs (24 non-terminal states) or by those and the ball-strike count (288 non-terminal states) \parencite{bukiet_markov_1997}. Our construction of the state extends them by further including the number of disengagements as a component of the state, thereby allowing us to analyze the game under the new pickoff limit rule.

    {\bf 2. Action space.} We limit the agency of runners as well as pitchers to states in which only the first base is occupied (and the other bases are unoccupied). Though restricting the agency to states in which only the first base is occupied (i.e., the component $b$ in the state variable equals $(1,0,0)$) limits the scope, the situations in which the runner might steal the second base comprises roughly $71\%$ of stolen base attempts in 2023. We emphasize that this  excludes the scenario with $b = (1, 0, 1)$ in which the first-base runner will often attempt a stolen base to lure a throw from the catcher, giving the third-base runner a chance to steal home. We leave this for future research.

    For states in which only the first base is occupied, the first-base runner's action is the lead distance chosen in feet from the first base. The pitcher has only two actions: attempt a pickoff or throw a pitch. Formally, the action space of the first-base runner, $A_{\text{R}}$, and the action space of the pitcher, $A_{\text{P}}$, are as follows.

    \begin{align}
    \label{eqn:runner_action_sapce}
        A_\text{R}(s) &\defeq \begin{cases}
            {\cal L} \text{ if } s = (b,c,d,o) \text{ and } b = (1,0,0)\\
            \{\delta\} \text{ if } s = (b,c,d,o) \text{ and } b \neq (1,0,0), \text{ or } s = \Delta, 
        \end{cases} \ \forall s \in {\cal S}\\
        \nonumber
        \text{ where } {\cal L} &\defeq [0,20] \text{ is the set of allowable lead distances for the first-base runner.}
    \end{align}    
    \begin{align}
    \label{eqn:pitcher_action_sapce}
        A_{\text{P}}(s,a_\text{R}) &\defeq \begin{cases}
            {\cal P} \text{ if } s = (b,c,d,o) \text{ and } b = (1,0,0) \\
            \{\delta\} \text{ if } s = (b,c,d,o) \text{ and } b \neq (1,0,0), \text{ or } s = \Delta,         \end{cases} \ \forall s \in {\cal S}, \ a_\text{R} \in A_\text{R}(s)\\
            \nonumber
        \text{ where } {\cal P} &\defeq \{0,1\}, \text{ with } p = 1 \text{ corresponding to a pickoff attempt and } 0 \text{  otherwise.}
    \end{align} 
    Here, $\delta$ is a dummy element that captures the lack of agency. We note that the state of our stochastic game progresses even when no actions are played by both the players. This will be modeled through appropriate transition probabilities.

    {\bf 3. Transition probabilities.} The uncertainty in our sequential game is modeled through appropriate transition probabilities. Basically, quantities describe the conditional probability of transitioning to a future state given the sequence of previously visited states and the actions played in those states. Our sequential game model assumes that the transition probabilities satisfy the Markov property, i.e., conditional on the entire history sequence, the probability of transitioning to a next state depends only on the current state and the actions played in that state.  Following the convention in the literature, we denote the transition probability as $p(\cdot|s,a_\text{R},a_\text{P}) \ \forall s \in {\cal S}, a_\text{R} \in A_\text{R}(s), a_\text{P} \in A_\text{P}(s,a_\text{R})$. Section \ref{sec:transition-probability-model} details our specific construction of transition probabilities. At a high level, we use generalized linear mixed-effects models to estimate runner outcome (e.g., unsuccessful pickoff attempt or successful stolen base attempt) probabilities which depend on lead distance and the pitcher's pickoff attempt decision. This along with appropriate conditional empirical frequencies will be used to construct the transition probabilities.
    % SP: We are deleting this sentence because we can't remember why we need it
    %Any state for which the number of baserunners plus the number of outs is at least three (i.e., $g(b) + o \ge 3$) may transition to the absorbing state with positive probability because there is always the possibility of recording enough outs on a single play to end the inning. \todo{Siva: This last sentence needs to go to the section on transition probabilities.}

    {\bf 4. Reward function.} The final component of our sequential game is the reward function, which models the payoff received by the players. Since we use a  zero-sum sequential game, the payoff obtained by one player comes at the expense of the other player. We choose the number of runs scored as the reward function. This design choice is more suitable for the early and middle innings of a game and less suitable for the later innings, particularly the bottom of the ninth or extra innings, when the offense knows how many runs they need to score to win the game. For any state $s \in {\cal S}$, the reward function captures the number of runs scored by the batting team in transitioning from state $s$ to $s'$ given the actions of the first-base runner and the pitcher. Formally,
    \begin{align}
      \label{eqn:runner_pitcher_reward}
        r(s'|s,a_\text{R},a_\text{P}) \defeq r(s'|s) \ \forall s,s' \in {\cal S}, a_\text{R} \in A_\text{R}(s), a_\text{P} \in A_\text{P}(s,a_\text{R}),
    \end{align}
    where
    \begin{align}
      \label{eqn:common_reward}
      r(s'|s) \defeq  \begin{cases}
          (g(b) + o) - (g(b') + o') + \mathbb{I}\{c' = (0, 0),\, d' = 0\} & \mbox{ if } s,\, s' \in \mathcal{X}\\
          \hfil s' & \text{ if } s \in \mathcal{X},\, s' \in \mathcal{P}\\
          \hfil 0 & \text{ if } s' = \Delta,
      \end{cases} 
    \end{align}
    where $g: \{0, 1\}^3 \rightarrow \{0, 1, 2, 3\}$ counts the number of runners on base. Note that our reward function in \eqref{eqn:runner_pitcher_reward} depends only on the current state and the next state, and not on the actions of the first-base runner or the pitcher. The formula in \eqref{eqn:common_reward} counts the total number of baserunners and outs before and after the transition and subtracts one if the transition corresponds to the end of a plate appearance. Since the number of runs scored on the last play can not be uniquely determined from the current state, we use the penultimate states to track the number of runs scored on the last play. We note that our transition probabilities will be constructed such that these penultimate states transition to the terminal end-of-inning state with probability $1$. Since the reward function \eqref{eqn:runner_pitcher_reward} depends only on the current and the next state, it looks like it is independent of the actions of the players. However, the next state to which our sequential game transitions is (probabilistically) determined by the actions of both the players through the transition probabilities, thereby making the reward function indirectly depend on the actions of both the players.

    Our sequential game models the progression of an inning as follows. Let $s \in {\cal S}$ denote the state of the sequential game from which a play begins. Upon observing this state, the first-base runner picks an action $a_\text{R} \in A_\text{R}(s)$. The pitcher observes the state $s$ and the first-base runner's action $a_\text{R}$, and then picks an action $a_\text{P} \in A_\text{P}(s,a_\text{R})$. After playing $a_\text{R}$ and $a_\text{P}$, the sequential game then transitions into state $s'$ with probability $p(s'|s,a_\text{R},a_\text{P})$ and the batting team collects a reward $r(s'|s,a_\text{R},a_\text{P})$ in the zero-sum game. A play then begins from $s'$ and the process repeats. The goal of the batting team is to maximize (through the actions of the first-base runner) the expected total rewards over the plays in the inning while the goal of the fielding team is to minimize (through the actions of the pitcher) the expected total rewards obtained by the batting team.

>>>>

\usepackage{amsmath, amssymb, amsthm, amsfonts, authblk, color, enumerate, framed, graphicx, mathrsfs, mathtools, mdframed, subcaption, tabularx, theoremref, titlesec, verbatim, xfrac, soul, mathtools, wrapfig, tikz, float}

@@

\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}
\newcommand{\defeq}{\stackrel{\text{def}}{=}}
\newcommand{\abs}[1]{\left\lvert#1\right\rvert}
\newcommand{\revision}[1]{{\color{orange} #1}}

\title{
  Lead distance under a pickoff limit in Major League Baseball:\\A sequential game model
}
\author[1]{Scott Powers}
\author[2]{Sivaramakrishnan Ramani}
\author[3]{Jacob Hahn}
\author[2]{Andrew J. Schaefer}
\affil[1]{Department of Sport Management, Rice University}
\affil[2]{Dept of Computational Applied Mathematics and Operations Research, Rice University}
\affil[3]{Department of Statistics, Rice University}

@@

  \begin{abstract}
    Recently, Major League Baseball (MLB) limited pitchers to three pickoff attempts, creating a cat-and-mouse game between pitcher and runner. Each failed attempt adds pressure on the pitcher to avoid using another, and the runner can intensify this pressure by extending their leadoff toward the next base. We model this dynamic as a two-player zero-sum sequential game in which the runner first chooses a lead distance, and then the pitcher chooses whether to attempt a pickoff. We establish optimality characterizations for the game and present variants of value iteration and policy iteration to solve the game. Using lead distance data, we estimate generalized linear mixed-effects models for pickoff and stolen base outcome probabilities given lead distance, context, and player skill. We compute the game-theoretic equilibria under the two-player model, as well as the optimal runner policy under a simplified one-player Markov decision process (MDP) model. In the one-player setting, our results establish an actionable rule of thumb: the Two-Foot Rule, which recommends that a runner increase their lead by two feet after each pickoff attempt.
  \end{abstract}

@@

    On May 17, 2025, the Chicago Cubs hosted the Chicago White Sox in the second game of their Crosstown Series. In the bottom of the first inning, White Sox pitcher Sean Burke faced Cubs batter Michael Busch with two outs and Cubs runner Kyle Tucker on first base. Before the 1-1 pitch, Burke pivoted and threw to first base behind Tucker, who retreated safely before a tag. Tucker had 10 stolen bases in 10 attempts on the young season, the most of any base stealer who had not been caught \parencite{fangraphs_major_2025}. On the White Sox broadcast, color commentator Steve Stone explained to the audience, ``The only problem is now there's two throws over there, and Tucker's gonna get a much bigger lead.'' On the next pitch, Tucker stole his 11$^{th}$ base of the season, the first Burke had allowed. ``I don't think anybody in the room could have thrown him out on that play,'' said Stone.

    For background, the stolen base is an exciting play in which a runner takes off on the pitch, attempting to reach the next base before the catcher can throw the to a fielder at the base to tag the runner out \parencite{mlb_baseball_2025}. Whether attempting a stolen base or not, the runner takes a {\it lead} from their base, commonly positioning themselves 8--12 feet away from their base, toward the next base. The longer the lead, the shorter the distance needed to run to reach the next base. The pitcher's primary deterrent against long leads is a {\it pickoff attempt}, in which they throw the ball to a fielder at the runner's base, attempting a tag before the runner returns safely to the base.

    MLB stolen base totals surged in the 1980s and 1990s but then waned in the 2000s and 2010s as teams cut back on the play, which generally requires a high success rate in order to have a positive impact on run scoring \parencite{tango_book_2007}. Before 2023, pickoff attempts could delay games because unsuccessful pickoff attempts do not advance the state of the game. Acknowledging this and other trends, MLB introduced several rule changes in 2023, with the goal of making the game more exciting \parencite{castrovince_pitch_2023}. With these changes, a pitch clock limits the time between pitches, and there are restrictions on infielder positioning, curtailing defensive shifts. Two changes are particularly relevant to stolen base attempts: Bases are now wider (reducing the distance between first and second base by 4.5 inches), and there are now limitations on pickoff attempts, or more formally, {\it disengagements}.

@@

    Section \ref{sec:sg-model} uses a two-payer zero-sum sequential game to model the evolving interaction between the runner and the pitcher.  Our model captures the vital aspects of the play while retaining analytical and computational tractability. Our model is closely related to the class of stochastic/dynamic Stackelberg games \parencite{goktas2022zero,li2017review,lopez2022stationary,vasal2020stochastic,vorobeychik2012computing}, which combines the elements of Stackelberg and stochastic games. The main difference between our infinite-horizon model and the stochastic Stackelberg games literature is that we do not discount the rewards. The rationale for this modeling choice is that our optimization objective---expected runs in the inning---by definition includes no discount. We establish a maximin Bellman equation for our sequential game and propose a variant of value iteration and policy iteration to determine the optimal strategies for runner and pitcher.

    In addition to developing a model which captures the important features of the pitcher-runner game, we analyze the model using recently available real-world MLB data. The most innovative component of our parameterization is the construction of transition probabilities. The outcome of each play is influenced by factors outside of our state variable, such as the sprint speed of the runner and the arm strength of the catcher. As a result, constructing transition probabilities using empirical conditional estimates (which use only the information from the current state and the actions of the player) is likely to be less accurate compared with a method that incorporates the player-specific factors.

    Section \ref{sec:transition-probability-model} uses a two-step procedure that includes the player-specific factors to estimate the transition probabilities. In the first step, we use a mixed-effects logistic regression model that calculates the probability of various runner outcomes taking into account the actions of the players and the player-specific information. Next, we construct an empirical probability distribution over the next state conditioned upon the current state, the runner outcomes, and the actions of the players. Finally, we determine the transition probability as the product of these two probabilities.

    Section \ref{sec:results} presents a numerical study on recent real-world data which illustrates that, relative to the optimal policy, observed pitcher and runner decisions are conservative. In current practice, pitchers attempt fewer pickoffs than optimal, and runners take shorter leads than optimal. Relative to observed behavior, optimal play would favor the offense slightly. Moreover, when restricting our model to a single-player setup which allows agency only for the runner, the results reveal an actionable recommendation: the Two-Foot Rule, which recommends that a runner increase their lead by two feet after each pickoff attempt.

@@

    Several studies in the sports economics literature have examined the extent to which professional athletes play minimax strategy. \textcite{palacios-huerta_professionals_2003} found evidence consistent with minimax equilibrium behavior in the mixed strategy game of the soccer penalty kick, in which the kicker chooses to shoot the ball left or right and the goalkeeper simultaneously chooses to dive left or right for the save. \textcite{palacios-huerta_experientia_2008} followed this with a laboratory experiment featuring professional soccer players and college students simulating the penalty kick decision using cards, showing that the professionals played minimax but the students did not. On the other hand, \textcite{kovash_professionals_2009} found evidence inconsistent with minimax equilibrium behavior in the National Football League (teams called more run plays and fewer pass plays than theoretically optimal) and in MLB (pitchers threw more fastballs than theoretically optimal).

    The history of quantitative analysis in baseball has deep roots in operations research (e.g., \cite{lindsey_investigation_1963}; \cite{freeze_analysis_1974}; \cite{pankin_evaluating_1978}; \cite{russell_devising_1994}; \cite{bukiet_markov_1997}). The Markov process is particularly well suited for baseball, where the game's highly structured progression lends it well to an intuitive state space model \parencite{sokol_intuitive_2004}. \textcite{hirotsu_markov_2003} used a Markov model to inform substitution decision-making (specifically, pinch hitting) in baseball. The Markov process is also popular across sport analytics. \textcite{hirotsu_using_2002} used a four-state Markov model for soccer to inform substitution decision-making based on how substitutions impact transition probabilities between states. Recently, \textcite{chan_points_2021} used a Markov model to evaluate team performance in football. More specifically, the MDP has become a particularly popular model in recent years for analyzing decision-making in sports, with applications in tennis \parencite{nadimpalli_when_2013}, baseball \parencite{hirotsu_using_2019}\todo{pull this out with its own sentence at the end}, basketball \parencite{sandholtz_markov_2020}, darts \parencite{baird_optimising_2020} and football \parencite{biro_reinforcement_2022}.

    An intermediate result of the present work is a probability model for stolen base attempts and successes (as well as an analysis of how those probabilities changed with the 2023 MLB rule changes), a topic which has not received much attention. \textcite{loughin_assessing_2008} estimated linear mixed-effects models for stolen base attempts and successes, using random effects for pitcher and catcher (not the runner, notably) and fixed effects for balls, strikes, outs, score, venue and pitcher hand. They showed that the spread of pitcher effects is slightly greater than the spread of catcher effects. That analysis did not use data on runner sprint speed and catcher arm strength. More recently, \textcite{stanley_modeling_2023} modeled stolen base success probability using both logistic regression and a random forest, including runner sprint speed and catcher arm strength as features but excluding player identities. That analysis was limited to data prior to the 2023 MLB rule changes that incentivized base stealing.

    \section{The two-player sequential game model}
    \label{sec:sg-model}
    In this section, we present our two-player zero-sum sequential game that models the dynamic between runner and pitcher under the pickoff limit rule. Though our sequential game describes only a portion of an inning, it captures the vital aspects that enable the generation of valuable and actionable insights for the runner and the pitcher. Before describing the specifics of our model, we note that our model assumes a representative pitcher and a representative runner for the whole inning, rather than capturing full lineups and substitutions of players with varying skills. Without this assumption, we would end up with a nonstationary (infinite-horizon) two-player sequential game, which appears to be significantly more challenging to analyze when compared with the stationary version. The four components of our sequential game are as follows.

    {\bf 1. State space.} We use $s=(b,c,d,o) \in {\cal X}$ to capture a play, where     
    % The mid-inning game state $s \in \mathcal{M}$ is given by $s = (b, c, d, o)$, where:
    \begin{align*}
      b &\in \{0, 1\}^3 \text{ represents which bases are occupied by runners};\\
      c &\in \{0, 1, 2, 3\} \times \{0, 1, 2\} \text{ represents the ball-strike count};\\
      d &\in \{0, 1, 2\} \text{ represents the number of disengagements already used; and}\\
      o &\in \{0, 1, 2\} \text{ represents the number of outs}.
    \end{align*}
    The number of elements in ${\cal X}$ equals $2^3 \times (4 \times 3) \times 3 \times 3 = 864$. We use $\mathcal P$ to denote the set of four penultimate states and $\Delta$ to denote the terminal end-of-inning state. The purpose of the penultimate states is to keep track of the number of runs that are scored on the transition to the terminal state. A penultimate state $s \in \mathcal{P} \equiv \{0, 1, 2, 3\}$ represents the number of runs scored on the final play of the inning. We use $\mathcal{S} \equiv \mathcal{X} \cup \mathcal{P} \cup \{\Delta\}$ to denote the set of all states and $S_t$ to denote the state at play $t$ in a (fixed) inning. Traditionally, the state is defined by the bases occupied and the number of outs (24 non-terminal states) or by those and the ball-strike count (288 non-terminal states) \parencite{bukiet_markov_1997}. Our construction of the state extends them by further including the number of disengagements as a component of the state, thereby allowing us to analyze the game under the new pickoff limit rule.

    {\bf 2. Action space.} We limit the agency of runners and pitchers to states in which only first base is occupied 
    (i.e., the component $b$ in the state variable equals $(1,0,0)$). Although this restriction limits the scope, these states covered roughly $71\%$ of stolen base attempts in 2023. We emphasize that this  excludes the scenario with $b = (1, 0, 1)$ in which the first-base runner will often attempt a stolen base to lure a throw from the catcher, giving the third-base runner a chance to steal home and establishing a different dynamic \parencite{miller_first-and-third_2024}. We leave this for future research.

    For states in which only first base is occupied, the first-base runner's action is the lead distance from first base, chosen in feet. The pitcher has only two actions: attempt a pickoff or throw a pitch. Formally, the action space of the first-base runner, $A_{\text{R}}$, and the action space of the pitcher, $A_{\text{P}}$, are as follows.

    \begin{align}
    \label{eqn:runner_action_sapce}
        A_\text{R}(s) &\defeq \begin{cases}
            {\cal L} \text{ if } s = (b,c,d,o) \text{ and } b = (1,0,0)\\
            \{\delta\} \text{ if } s = (b,c,d,o) \text{ and } b \neq (1,0,0), \text{ or } s \in {\cal P} \cup \{\Delta\}
        \end{cases} \ \forall s \in {\cal S}\\
        \nonumber
        \text{ where } {\cal L} &\defeq [0,20] \text{ is the set of allowable lead distances for the first-base runner.}
    \end{align}
    \begin{align}
    \label{eqn:pitcher_action_sapce}
        A_{\text{P}}(s,a_\text{R}) &\defeq \begin{cases}
            {\cal P} \text{ if } s = (b,c,d,o) \text{ and } b = (1,0,0) \\
            \{\delta\} \text{ if } s = (b,c,d,o) \text{ and } b \neq (1,0,0), \text{ or } s \in {\cal P} {\cup} \{\Delta\},         \end{cases} \ \forall s \in {\cal S}, \ a_\text{R} \in A_\text{R}(s)\\
            \nonumber
        \text{ where } {\cal P} &\defeq \{0,1\}, \text{ with } p = 1 \text{ corresponding to a pickoff attempt and } 0 \text{  otherwise.}
    \end{align} 
    Here, $\delta$ is a dummy element that captures the lack of agency. We note that the state of our sequential game progresses even when no actions are played by both players. We will model this through appropriate transition probabilities.

    {\bf 3. Transition probabilities.} The uncertainty in our sequential game is modeled through transition probabilities, i.e., the conditional probability of transitioning to a future state given the sequence of previously visited states and the actions played in those states. Our sequential game assumes that the transition probabilities satisfy the Markov property.  Following the convention in the literature, we denote the transition probability as $p(\cdot|s,a_\text{R},a_\text{P}) \ \forall s \in {\cal S}, a_\text{R} \in A_\text{R}(s), a_\text{P} \in A_\text{P}(s,a_\text{R})$. Section \ref{sec:transition-probability-model} details our specific construction of transition probabilities, which uses generalized linear mixed-effects models to estimate runner outcome (e.g., unsuccessful pickoff attempt or successful stolen base attempt) probabilities which depend on lead distance and the pitcher's pickoff decision, as well as player-specific factors. These, along with appropriate conditional empirical frequencies, will be used to construct the transition probabilities.
    % SP: We are deleting this sentence because we can't remember why we need it
    %Any state for which the number of baserunners plus the number of outs is at least three (i.e., $g(b) + o \ge 3$) may transition to the absorbing state with positive probability because there is always the possibility of recording enough outs on a single play to end the inning. \todo{Siva: This last sentence needs to go to the section on transition probabilities.}

    {\bf 4. Reward function.} The final component of our sequential game is the reward function, which models the payoff received by the players. Since we use a  zero-sum sequential game, the payoff obtained by one player comes at the expense of the other player. We choose the number of runs scored as the reward function. This design choice is more suitable for the early and middle innings of a game and less suitable for the later innings, particularly the bottom of the ninth or extra innings, when the offense knows how many runs they need to score to win the game. For any state $s \in {\cal S}$, the reward function captures the number of runs scored by the batting team in transitioning from state $s$ to $s'$ given the actions of the first-base runner and the pitcher. Formally,
    \begin{align}
      \label{eqn:runner_pitcher_reward}
        r(s'|s,a_\text{R},a_\text{P}) \defeq r(s'|s) \ \forall s,s' \in {\cal S}, a_\text{R} \in A_\text{R}(s), a_\text{P} \in A_\text{P}(s,a_\text{R}),
    \end{align}
    where
    \begin{align}
      \label{eqn:common_reward}
      r(s'|s) \defeq  \begin{cases}
          (g(b) + o) - (g(b') + o') + \mathbb{I}\{c' = (0, 0),\, d' = 0\} & \mbox{ if } s,\, s' \in \mathcal{X}\\
          \hfil s' & \text{ if } s \in \mathcal{X},\, s' \in \mathcal{P}\\
          \hfil 0 & \text{ if } s' = \Delta,
      \end{cases} 
    \end{align}
    where $g: \{0, 1\}^3 \rightarrow \{0, 1, 2, 3\}$ counts the number of runners on base. Note that our reward function in \eqref{eqn:runner_pitcher_reward} depends only on the current state and the next state, and not on the actions of the first-base runner or the pitcher. The formula in \eqref{eqn:common_reward} infers the number of runs scored by counting the offensive players captured by the states before and after the transition. If the transition corresponds to the end of a plate appearance (i.e. $c' = (0, 0),\, d' = 0$), then we account for one more offensive player (the batter) in the next state. Since the number of runs scored on the last play cannot be uniquely determined from the current state, we use the penultimate states to track the number of runs scored on the last play. We note that our transition probabilities will be constructed such that these penultimate states transition to the terminal end-of-inning state with probability $1$. Since the reward function \eqref{eqn:runner_pitcher_reward} depends only on the current and the next state, it appears to be independent of the actions of the players. However, the next state to which our sequential game transitions is (probabilistically) determined by the actions of both the players through the transition probabilities, thereby making the rewards depend indirectly on the actions of both players.

    Our sequential game models the progression of an inning as follows. Let $s \in {\cal S}$ denote the state of the sequential game from which a play begins. Upon observing this state, the first-base runner picks an action $a_\text{R} \in A_\text{R}(s)$. The pitcher observes the state $s$ and the first-base runner's action $a_\text{R}$, and then picks an action $a_\text{P} \in A_\text{P}(s,a_\text{R})$. After playing $a_\text{R}$ and $a_\text{P}$, the sequential game then transitions into state $s'$ with probability $p(s'|s,a_\text{R},a_\text{P})$ and the batting team collects a reward $r(s'|s,a_\text{R},a_\text{P})$ in the zero-sum game. A play then begins from $s'$, and the process repeats. The goal of the batting team is to maximize (through the actions of the first-base runner) the expected total rewards over the plays in the inning while the goal of the fielding team is to minimize (through the actions of the pitcher) the expected total rewards obtained by the batting team.

    Intuitively, any sound strategy for the first-base runner and the pitcher depends on the state of the sequential game (for the pitcher, it also depends on the actions played by the first-base runner). This is formally captured using the concept of policy. A (deterministic, stationary) policy for the first-base runner is a decision-rule that prescribes the action to be played in every possible state. Similarly, a (deterministic, stationary) policy for the pitcher is a decision-rule that prescribes the action to be played for every possible state and the corresponding actions of the first-base runner in those states. A mathematical treatment of policies is presented in the Appendix. We denote the set of policies for the runner and pitcher by $\Pi_\text{R}$ and $\Pi_\text{P}$, respectively.
